---
title: "A practical guide to ggplot2 and ggdist"
---

::: {.callout-note}
This report was generated using AI under general human direction. At the time of generation, the contents have not been comprehensively reviewed by a human analyst.

<!--
To indicate human review: Delete the line above about contents not being reviewed, and replace this comment with:
The contents have been reviewed and validated by [Your Name], [Your Role] on [Date].
-->
:::

# Overview

This guide focuses on reusable plotting patterns for `ggplot2` and `ggdist`. The examples use built-in R datasets so the document can render without external data files.

The code below is written to be modular: the helper functions can be copied into other scripts, packages, Shiny apps, or reports with minimal modification.

# Setup

The next chunk loads the packages used throughout the guide. The examples rely on `ggplot2` for general plotting, `ggdist` for distribution and interval geoms, and `dplyr` for light data preparation.

```{r}
#| label: load-packages
library(ggplot2)
library(ggdist)
library(dplyr)
```

# Reusable plotting helpers

The helpers in this section are designed for repeated use across projects. They take a data frame and column names, which makes them easy to drop into a function, app, or analysis script.

## A publication-friendly theme

Use this when you want a light, readable theme for exploratory graphics or report-ready figures. It works well for scatter plots, line charts, and summaries built from small or medium-sized tabular data.

```{r}
#| label: define-theme-helper
#| echo: true

plot_theme_publication <- function(base_size = 12, base_family = "") {
  theme_minimal(base_size = base_size, base_family = base_family) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title.position = "plot",
      plot.title = element_text(face = "bold"),
      axis.title = element_text(face = "bold"),
      legend.position = "bottom"
    )
}
```

## A reusable scatter plot template

Use this for two continuous variables, optionally coloured by a grouping variable. This pattern is common in customer analysis, quality control, ecological data, and any tabular dataset with paired measurements.

```{r}
#| label: define-scatter-helper
#| echo: true

plot_scatter_template <- function(data, x, y, color = NULL, smooth = TRUE) {
  p <- ggplot(data, aes(x = {{ x }}, y = {{ y }})) +
    geom_point(alpha = 0.75, size = 2)

  if (!rlang::quo_is_null(rlang::enquo(color))) {
    p <- p + aes(color = {{ color }})
  }

  if (smooth) {
    p <- p + geom_smooth(se = FALSE, linewidth = 0.8)
  }

  p +
    labs(x = NULL, y = NULL) +
    plot_theme_publication()
}
```

## A reusable time-series line template

Use this when the x-axis is ordered data such as dates, time stamps, or sequence numbers. It is useful for operational metrics, environmental data, or any repeated measurement over time.

```{r}
#| label: define-line-helper
#| echo: true

plot_line_template <- function(data, x, y, color = NULL) {
  p <- ggplot(data, aes(x = {{ x }}, y = {{ y }})) +
    geom_line(linewidth = 0.8)

  if (!rlang::quo_is_null(rlang::enquo(color))) {
    p <- p + aes(color = {{ color }})
  }

  p +
    labs(x = NULL, y = NULL) +
    plot_theme_publication()
}
```

## A reusable summary interval template

Use this when you want to show a point estimate and uncertainty interval for groups. This is especially helpful for survey data, experiment summaries, and any grouped numeric outcome.

```{r}
#| label: define-interval-helper
#| echo: true

plot_pointinterval_template <- function(data, x, y) {
  ggplot(data, aes(x = {{ x }}, y = {{ y }})) +
    stat_pointinterval(
      point_interval = median_qi,
      .width = c(0.5, 0.8, 0.95)
    ) +
    labs(x = NULL, y = NULL) +
    plot_theme_publication()
}
```

## A reusable distribution template with ggdist

Use this for showing a full distribution, not just a summary. It works well for skewed variables, multi-modal data, and group comparisons where the shape of the distribution matters.

```{r}
#| label: define-halfeye-helper
#| echo: true

plot_halfeye_template <- function(data, x, group) {
  ggplot(data, aes(x = {{ x }}, y = {{ group }})) +
    stat_halfeye(
      adjust = 0.8,
      .width = c(0.5, 0.8, 0.95)
    ) +
    labs(x = NULL, y = NULL) +
    plot_theme_publication()
}
```

# Examples with built-in R datasets

## Scatter plot with `iris`

Use this example when you have two continuous measurements and a categorical grouping variable. The `iris` dataset is a classic example because it contains flower measurements by species.

```{r}
#| label: iris-scatter-example
plot_scatter_template(
  data = iris,
  x = Sepal.Length,
  y = Petal.Length,
  color = Species,
  smooth = FALSE
) +
  labs(
    title = "Sepal length and petal length in iris",
    subtitle = "Points are coloured by species"
  )
```

## Time-series line plot with `airquality`

Use this example when your data are ordered over time or another sequence. The `airquality` dataset is built into R and works well for showing a daily trend after a small amount of date preparation.

```{r}
#| label: airquality-line-example
airquality_dates <- airquality |>
  mutate(Date = as.Date(sprintf("1973-%02d-%02d", Month, Day)))

plot_line_template(
  data = airquality_dates |> filter(!is.na(Temp)),
  x = Date,
  y = Temp
) +
  labs(
    title = "Daily temperature in New York City, 1973",
    subtitle = "Built from the native airquality dataset"
  )
```

## Distribution and interval summary with `mtcars`

Use this example when you want to compare a numeric outcome across groups. The `mtcars` dataset is a convenient built-in dataset for grouped summaries.

```{r}
#| label: mtcars-interval-example
mtcars |>
  mutate(cyl = factor(cyl)) |>
  plot_pointinterval_template(x = cyl, y = mpg) +
  labs(
    title = "Fuel efficiency by cylinder count",
    subtitle = "Point estimates with median and quantile intervals"
  )
```

## Distribution display with `ToothGrowth`

Use this example when the distribution itself matters, not just the mean. `ToothGrowth` is useful for showing the full spread of a numeric variable across treatment groups.

```{r}
#| label: toothgrowth-halfeye-example
ToothGrowth |>
  mutate(dose = factor(dose)) |>
  plot_halfeye_template(x = len, group = supp) +
  facet_wrap(~ dose) +
  labs(
    title = "Tooth length by supplement type and dose",
    subtitle = "Half-eye plots show shape and uncertainty together",
    x = "Tooth length",
    y = "Supplement"
  )
```

# Copy-paste patterns

The snippets below are the most directly reusable parts of the guide.

```{r}
#| label: reusable-snippets
#| echo: true

# Theme
plot_theme_publication <- function(base_size = 12, base_family = "") {
  theme_minimal(base_size = base_size, base_family = base_family) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title.position = "plot",
      plot.title = element_text(face = "bold"),
      axis.title = element_text(face = "bold"),
      legend.position = "bottom"
    )
}

# Scatter template
plot_scatter_template <- function(data, x, y, color = NULL, smooth = TRUE) {
  p <- ggplot(data, aes(x = {{ x }}, y = {{ y }})) +
    geom_point(alpha = 0.75, size = 2)

  if (!rlang::quo_is_null(rlang::enquo(color))) {
    p <- p + aes(color = {{ color }})
  }

  if (smooth) {
    p <- p + geom_smooth(se = FALSE, linewidth = 0.8)
  }

  p + labs(x = NULL, y = NULL) + plot_theme_publication()
}

# Interval template
plot_pointinterval_template <- function(data, x, y) {
  ggplot(data, aes(x = {{ x }}, y = {{ y }})) +
    stat_pointinterval(
      point_interval = median_qi,
      .width = c(0.5, 0.8, 0.95)
    ) +
    labs(x = NULL, y = NULL) +
    plot_theme_publication()
}
```

# Closing notes

For quick exploratory plots, start with the scatter and line templates from `ggplot2`. When you want to communicate uncertainty or distribution shape, switch to `ggdist` geoms such as `stat_pointinterval()` or `stat_halfeye()`.

The main design principle behind the code in this guide is consistency: keep the theme, axis handling, and helper signatures similar so the same snippets can move between notebooks, reports, and apps.
